---
title: "Part 11"
output:
  BiocStyle::html_document
---

```{r, message=FALSE, include=!exists(".standalone"), eval=!exists(".standalone")}
library("BloodCancerMultiOmics2017")
library("Biobase")
library("SummarizedExperiment")
library("AnnotationDbi")
library("org.Hs.eg.db")
library("dplyr")
library("abind")
library("reshape2")
library("RColorBrewer")
library("glmnet")
library("ipflasso")
library("ggplot2")
library("grid")
library("DESeq2")
```

```{r echo=FALSE}
plotDir = ifelse(exists(".standalone"), "", "part11/")
if(plotDir!="") if(!file.exists(plotDir)) dir.create(plotDir)
```

```{r}
options(stringsAsFactors=FALSE)
```


# Drug response prediction

Drug response heterogeneity is caused by the unique deregulations in biology of the tumor cell. Those deregulations leave trace on the different molecular levels and have a various impact on cell's drug sensitivity profile. Here we use multivariate regression to integrate information from the multi-omic data in order to predict drug response profiles of the CLL samples.

Loading the data.
```{r}
data(list=c("conctab", "drpar", "drugs", "patmeta", "lpdAll", "dds", "mutCOM",
"methData"))
```


## Assesment of omics capacity in explaining drug response

### Data pre-processing

Filtering steps and transformations.
```{r}
e<-dds
colnames(e)<-colData(e)$PatID


#only consider CLL patients
CLLPatients<-rownames(patmeta)[which(patmeta$Diagnosis=="CLL")]

#Methylation Data
methData = t(assay(methData)) 


#RNA Data
eCLL<-e[,colnames(e) %in% CLLPatients]
###
#filter out genes without gene namce
AnnotationDF<-data.frame(EnsembleId=rownames(eCLL),stringsAsFactors=FALSE)
AnnotationDF$symbol <- mapIds(org.Hs.eg.db,
                     keys=rownames(eCLL),
                     column="SYMBOL",
                     keytype="ENSEMBL",
                     multiVals="first")
eCLL<-eCLL[AnnotationDF$EnsembleId[!is.na(AnnotationDF$symbol)],]

#filter out low count genes
###
minrs <- 100
rs  <- rowSums(assay(eCLL))
eCLL<-eCLL[ rs >= minrs, ]
#variance stabilize the data
#(includes normalizing for library size and dispsersion estimation) 
vstCounts<-varianceStabilizingTransformation(eCLL)
vstCounts<-assay(vstCounts)
#no NAs in data
any(is.na(vstCounts))

#filter out low variable genes
ntop<-5000
vstCountsFiltered<-vstCounts[order(apply(vstCounts, 1, var, na.rm=T),
                                   decreasing = T)[1:ntop],]
eData<-t(vstCountsFiltered)
#no NAs
any(is.na(eData))

#genetics
#remove features with less than 5 occurences
mutCOMbinary<-channel(mutCOM, "binary")
mutCOMbinary<-mutCOMbinary[featureNames(mutCOMbinary) %in% CLLPatients,]
genData<-exprs(mutCOMbinary)
idx <- which(colnames(genData) %in% c("del13q14_bi", "del13q14_mono"))
genData <- genData[,-idx]
colnames(genData)[which(colnames(genData)=="del13q14_any")] = "del13q14"
minObs <- 5
#remove feutes with less than 5 occurecnes
genData<-genData[,colSums(genData, na.rm=T)>=minObs]

#IGHV
translation <- c(`U` = 0, `M` = 1)
stopifnot(all(patmeta$IGHV %in% c("U","M", NA)))
IGHVData <- matrix(translation[patmeta$IGHV], 
                   dimnames = list(rownames(patmeta), "IGHV"), ncol = 1)
IGHVData<-IGHVData[rownames(IGHVData) %in% CLLPatients,,drop=F]
#remove patiente with NA IGHV status
IGHVData<-IGHVData[!is.na(IGHVData), ,drop=F]
any(is.na(IGHVData))

#demographics (age and sex)
patmeta<-subset(patmeta, Diagnosis=="CLL")
gender <- ifelse(patmeta[,"Gender"]=="m",0,1)


# impute missing values in age by mean
ImputeByMean <- function(x) {x[is.na(x)] <- mean(x, na.rm=TRUE); return(x)}
age<-ImputeByMean(patmeta[,"Age4Main"])


demogrData <- cbind(age=age,gender=gender)
rownames(demogrData) <- rownames(patmeta)

#Pretreatment
pretreated<-patmeta[,"IC50beforeTreatment", drop=FALSE]

##### drug viabilites
summaries <- c(paste("viaraw", 1:5, sep=".") %>% `names<-`(paste(1:5)), 
               `4:5` = "viaraw.4_5", `1:5` = "viaraw.1_5")
a <- do.call( abind, c( along=3, lapply( summaries, 
                                         function(x) assayData(drpar)[[x]]))) 
dimnames(a)[[3]] <- names(summaries)
names(dimnames(a)) <- c( "drug", "patient", "summary" )
viabData <- acast( melt(a), patient ~ drug + summary )
rownames(viabData)<-c(substr(rownames(viabData),1,4)[1:3],
                      substr(rownames(viabData),1,5)[4:nrow(viabData)])
```


Check overlap of data and take care of missing values present in methylation and genetic data.
```{r}
# common patients 
Xlist<-list(RNA=eData, meth=methData, gen=genData, IGHV=IGHVData,
            demographics=demogrData, drugs=viabData, pretreated=pretreated)
PatientsPerOmic<-lapply(Xlist, rownames)
sapply(PatientsPerOmic, length)

allPatients<-Reduce(union, PatientsPerOmic)
PatientOverview<-sapply(Xlist, function(M) allPatients %in% rownames(M))
Patients <- (1:nrow(PatientOverview))
Omics <- (1:ncol(PatientOverview))
image(Patients,Omics, PatientOverview*1, axes=F, col=c("white", "black"),
      main="Sample overview across omics")
axis(2, at = 1:ncol(PatientOverview), labels=colnames(PatientOverview), tick=F)

commonPatients<-Reduce(intersect, PatientsPerOmic)
length(commonPatients)
XlistCommon<-lapply(Xlist, function(data) data[commonPatients,, drop=F])

#Take care of missing values (present in  genetic data)
ImputeByMean <- function(x) {x[is.na(x)] <- mean(x, na.rm=TRUE); return(x)}

#NAs in genetic
#remove feauters with less 90% completeness
RarlyMeasuredFeautres<-
  which(colSums(is.na(XlistCommon$gen))>0.1*nrow(XlistCommon$gen))
XlistCommon$gen<-XlistCommon$gen[,-RarlyMeasuredFeautres]
#remove patients with less than 90% of genetic feautres measured
IncompletePatients<-
  rownames(XlistCommon$gen)[
    (rowSums(is.na(XlistCommon$gen))>0.1*ncol(XlistCommon$gen))]
commonPatients<-commonPatients[!commonPatients %in% IncompletePatients]
XlistCommon<-lapply(XlistCommon, function(data) data[commonPatients,, drop=F])
#replace remaining NA by mean and round to 0 or 1
XlistCommon$gen<-round(apply(XlistCommon$gen, 2, ImputeByMean))

#NAs in methylation
#remove feauters with less 90% completeness
XlistCommon$meth<-
  XlistCommon$meth[,colSums(is.na(XlistCommon$meth))<0.1*nrow(methData)]
#impute remainin missing values by mean for each feautre across patients
XlistCommon$meth<-(apply(XlistCommon$meth, 2, ImputeByMean))

#final dimensions of the data
sapply(XlistCommon, dim)
```


Use top 20 PCs of methylation and expression as predictors.
```{r, fig.width=12, fig.height=10}
pcaMeth<-prcomp(XlistCommon$meth, center=T, scale. = F)
XlistCommon$MethPCs<-pcaMeth$x[,1:20]
colnames(XlistCommon$MethPCs)<-
  paste("meth",colnames(XlistCommon$MethPCs), sep="")

pcaExpr<-prcomp(XlistCommon$RNA, center=T, scale. = F)
XlistCommon$RNAPCs<-pcaExpr$x[,1:20]
colnames(XlistCommon$RNAPCs)<-paste("RNA",colnames(XlistCommon$RNAPCs), sep="")
```


Choose drug viabilites of interest as response variables.
```{r}
DOI <- c("D_006_1:5", "D_010_1:5", "D_159_1:5","D_002_4:5", "D_003_4:5",
         "D_012_4:5", "D_063_4:5", "D_166_4:5")
drugviab<-XlistCommon$drugs
drugviab<-drugviab[,DOI, drop=F]
colnames(drugviab) <- drugs[substr(colnames(drugviab),1,5),"name"]
```


Construct list of designs used and scale all predictors to mean zero and unit variance.
```{r}
ZPCs<-list(expression=XlistCommon$RNAPCs,
        genetic=XlistCommon$gen, 
        methylation= XlistCommon$MethPCs,
        demographics=XlistCommon$demographics, 
        IGHV=XlistCommon$IGHV,
        pretreated = XlistCommon$pretreated)
ZPCs$all<-do.call(cbind, ZPCs)
ZPCsunscaled<-ZPCs
ZPCsscaled<-lapply(ZPCs, scale)
lapply(ZPCsscaled, colnames)
```

Define colors.
```{r}
set1 <- brewer.pal(9,"Set1")
colMod<-c(paste(set1[c(4,1,5,3,2,7)],"88",sep=""), "grey")
names(colMod) <-
  c("demographics", "genetic", "IGHV","expression", "methylation", "pretreated",
    "all")
```


### Lasso using multi-omic data

Fit a linear model using Lasso explaining drug response by each one of the omic set separately as well as all together. As measure of explained variance use  R2 from linear models for unpenalized models (IGHV)
and fraction of variance explained, i.e. 1- cross-validated mean squared error/total sum of squares for others.

To ensure fair treatment of all features they are standardized to mean 0 and unit variance.
To study robustness the cross-validation is repeated 100-times to obtain the mean and standard deviation shown in the figure.

```{r, echo=F}
#Function to calculate Var Explained for Penalized Regression
R2ForPenRegPlusViz<-function(Z, drugviab, nfolds=10, alpha=1, nrep=100,
                      Parmfrow=c(2,4), ylimMax=0.4, standardize=TRUE){
Zlocal<-Z

set.seed(1030)
seeds<-sample(1:10000000, nrep)

RepeatedR2list<-lapply(1:nrep, function(outer){
#Use same folds for all omics to make comparable
set.seed(seeds[outer])
foldsAssignment<-sample(rep(seq(nfolds), length=nrow(drugviab)))

R2echOmicadj<-sapply(colnames(drugviab), function(dname) {
  d<-drugviab[,dname]
  sapply(names(Zlocal), function(nameZ) {
    pred<-Zlocal[[nameZ]]
    #fit a lasso model for omics with more than one features
    if(ncol(pred)>1){
    fitcv<-cv.glmnet(pred,d,family="gaussian", standardize=standardize,
                      alpha=alpha, foldid=foldsAssignment)
    R2 <- 1-min(fitcv$cvm)/fitcv$cvm[1]
    
    }else {#fit a liner model for single feautres (IGHV)
      fitlm<-lm(d~., data.frame(pred))
      R2<- summary(fitlm)$r.squared
    }
    R2
  })
}
  )
})
RepeatedR2<-RepeatedR2list
#calculate mean and sd across repitions wiht different folds
meanR2<-apply(simplify2array(RepeatedR2), 1:2, mean)
sdR2<-apply(simplify2array(RepeatedR2), 1:2, sd)

par(mfrow=Parmfrow, mar=c(7,5,3,3))
for (i in 1: ncol(meanR2)) {
    barc<-barplot(meanR2[,i], main= colnames(meanR2)[i], ylim=c(0,ylimMax),
                  las=2, col=colMod[rownames(meanR2)], ylab="R2")
    segments(barc, meanR2[,i]-sdR2[,i],barc, meanR2[,i]+sdR2[,i])
}
RepeatedR2
}
```

Fit model and show resulting omic-prediction profiles.
```{r lasso_main, echo=FALSE, fig.path=plotDir, dev=c("png", "pdf"), fig.width=9, fig.height=8}
#FIG# 5A
resultLassoPCA<-R2ForPenRegPlusViz(ZPCsscaled, drugviab, nfold=10, alpha=1,
                                   nrep=100, ylimMax=0.4)

df_resultLassoPCA <- melt(resultLassoPCA)
colnames(df_resultLassoPCA) <- c("omic", "drug", "R2", "run")
summaryR2 <- df_resultLassoPCA %>% group_by(omic, drug) %>% 
  dplyr::summarise(meanR2=mean(R2),sdR2 = sd(R2), nR2 = length(R2)) %>%
  mutate(seR2 = sdR2/sqrt(nR2))

ggplot(summaryR2, aes(x=omic, y=meanR2, fill=omic, group=omic))+
    geom_bar(stat = "identity") +  scale_fill_manual(values=colMod) + 
    geom_errorbar(aes(ymax = meanR2 + sdR2,ymin = meanR2 - sdR2), position = "dodge", width = 0.25) +facet_wrap(~drug, ncol=4) +theme_bw(base_size = 18) +ylab(bquote(R^2)) + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), axis.text = element_text(colour = "black")) +xlab("")+guides(fill=guide_legend(title="Data type")) 
```


Optionally, the model can also be fit for all drugs in the study.
```{r, eval=FALSE, fig.path='supp_'}
nfolds<-10
nrep<-100

DOI <-
  grepl("1:5",colnames(XlistCommon$drugs)) |
  grepl("4:5",colnames(XlistCommon$drugs))
drugviabAll<-XlistCommon$drugs
drugviabAll<-drugviabAll[,DOI]
colnames(drugviabAll) <- 
  paste0(drugs[substr(colnames(drugviabAll),1,5),"name"],
         substr(colnames(drugviabAll),6,9))

R2ForPenReg(Zscaled, drugviabAll, nfold=nfolds, alpha=1, nrep=nrep,
            Parmfrow=c(4,4), ylimMax=0.6)
```


## Lasso for drugs in a genetic-focussed model

We perform regression analysis by adaptive LASSO using genetic features, IGHV status (coded as 0 -1 for mutated/unmutated), pretreatment status (coded as 0 -1) and methylation cluster (coded as 0 for lowly-programmed (LP), 0.5 for intermediately-programmed (IP) and 1 for highly-programmed (HP)). For each model we select the optimal penalization parameter of the second step Lasso fit of the adaptive Lasso by repeated cross-validation to get robust results.

As output bar plots showing the coefficients of the selected predictors are produced.


### General definitions

We use the following abbreviations for the different data types.

|   | data type           |
|---|---------------------|
| M | methylation cluster |
| G | mutations           |
| V | viability           |
| I | ighv                |
| P | pretreatment        |



```{r echo=FALSE}
dataType = c(M="Methylation_Cluster", V="viab", G="gen", I="IGHV", P="pretreat")
```

Color definitions for the groups.
```{r echo=FALSE}
coldef<-list()
coldef["I"]<-brewer.pal(9, "Blues")[7]
coldef["M"]<-list(brewer.pal(9, "Blues")[c(1, 5, 9)])
coldef["G"]<-brewer.pal(8, "YlOrRd")[8]
coldef["P"]<-"chocolate4"
```


### Data pre-processing

Subselect CLL patients.
```{r}
lpdCLL = lpdAll[ , lpdAll$Diagnosis=="CLL"]
```

Prepare the data and limit the number of features by subselecting only those which include at least 5 recorded incidences. List the predictors.
```{r}
lpdCLL = lpdAll[ , lpdAll$Diagnosis=="CLL"]
lpdCLL = BloodCancerMultiOmics2017:::prepareLPD(lpd=lpdCLL, minNumSamplesPerGroup=5)
(predictorList = BloodCancerMultiOmics2017:::makeListOfPredictors(lpdCLL))
```

### Drug response prediction

The prediction will be made for the following drugs and concentrations.

|       | drug name     | conc |
|-------|---------------|------|
| D_159 | doxorubicine  | 1-5  |
| D_006 | fludarabine   | 1-5  |
| D_010 | nutlin-3      | 1-5  |
| D_166 | PRT062607 HCl | 4-5  |
| D_003 | idelalisib    | 4-5  |
| D_002 | ibrutinib     | 4-5  |
| D_012 | selumetinib   | 4-5  |
| D_063 | everolimus    | 4-5  |



```{r}
drs = list("1:5"=c("D_006", "D_010", "D_159"),
           "4:5"=c("D_002", "D_003", "D_012", "D_063", "D_166"))
```

```{r}
predvar = unlist(BloodCancerMultiOmics2017:::makePredictions(drs=drs,
                                 lpd=lpdCLL,
                                 predictorList=predictorList,
                                 lim=0.15, std=FALSE, adaLasso=TRUE),
                 recursive=FALSE)
```


```{r echo=FALSE}
details = function(dr, what) {
  predvar[[dr]][["plot"]][[what]]
}
```

```{r prediction-D_006, echo=FALSE, fig.path=plotDir, dev=c("png", "pdf"), fig.width=details("D_006","width"), fig.height=details("D_006","height"), eval=TRUE}
#FIG# 5B
grid.draw(details("D_006","plot"))
```

```{r prediction-D_010, echo=FALSE, fig.path=plotDir, dev=c("png", "pdf"), fig.width=details("D_010","width"), fig.height=details("D_010","height"), eval=TRUE}
#FIG# 5B
grid.draw(details("D_010","plot"))
```

```{r prediction-D_159, echo=FALSE, fig.path=plotDir, dev=c("png", "pdf"), fig.width=details("D_159","width"), fig.height=details("D_159","height"), eval=TRUE}
#FIG# 5B
grid.draw(details("D_159","plot"))
```

```{r prediction-D_002, echo=FALSE, fig.path=plotDir, dev=c("png", "pdf"), fig.width=details("D_002","width"), fig.height=details("D_002","height"), eval=TRUE}
#FIG# 5B
grid.draw(details("D_002","plot"))
```

```{r prediction-D_003, echo=FALSE, fig.path=plotDir, dev=c("png", "pdf"), fig.width=details("D_003","width"), fig.height=details("D_003","height"), eval=TRUE}
#FIG# 5B
grid.draw(details("D_003","plot"))
```

```{r prediction-D_012, echo=FALSE, fig.path=plotDir, dev=c("png", "pdf"), fig.width=details("D_012","width"), fig.height=details("D_012","height"), eval=TRUE}
#FIG# 5B
grid.draw(details("D_012","plot"))
```

```{r prediction-D_063, echo=FALSE, fig.path=plotDir, dev=c("png", "pdf"), fig.width=details("D_063","width"), fig.height=details("D_063","height"), eval=TRUE}
#FIG# 5B
grid.draw(details("D_063","plot"))
```

```{r prediction-D_166, echo=FALSE, fig.path=plotDir, dev=c("png", "pdf"), fig.width=details("D_166","width"), fig.height=details("D_166","height"), eval=TRUE}
#FIG# 5B
grid.draw(details("D_166","plot"))
```

Plot the legends.
```{r}
legends = BloodCancerMultiOmics2017:::makeLegends(legendFor=c("G","I","M", "P"))
```

```{r legend, echo=FALSE, fig.path=plotDir, dev=c("png", "pdf"), fig.width=legends[["width"]], fig.height=legends[["height"]]}
#FIG# 5B legend
grid.draw(legends[["plot"]])
```

Additionaly we plot the prediction for rotenone.
```{r}
drs_rot = list("4:5"=c("D_067"))
predvar_rot = unlist(BloodCancerMultiOmics2017:::makePredictions(drs=drs_rot,
                                 lpd=lpdCLL,
                                 predictorList=predictorList,
                                 lim=0.23, std=FALSE, adaLasso=TRUE),
                 recursive=FALSE)
```

```{r prediction-D_067, echo=FALSE, fig.path=plotDir, dev=c("png", "pdf"), fig.width=predvar_rot[["D_067"]][["plot"]][["width"]], fig.height=predvar_rot[["D_067"]][["plot"]][["height"]], eval=TRUE}
#FIG# S26
grid.draw(predvar_rot[["D_067"]][["plot"]][["plot"]])
```


In a same way the prediction for all the drugs can be made.
```{r}
alldrs = unique(fData(lpdCLL)[fData(lpdCLL)$type=="viab","id"])
drs = list("1:5"=alldrs, "4:5"=alldrs)
```

```{r, eval=TRUE}
predvar2 = BloodCancerMultiOmics2017:::makePredictions(drs=drs,
                                        lpd=lpdCLL,
                                        predictorList=predictorList,
                                        lim=0.23)
```

### Effect of pre-treatment

In order to find out the effect of pre-treatment on the predictions of drug response we provide the following overview. The summary is made for all drugs, separating however, on ranges of drug concentrations (mean drug effect of: all five (1-5) and two lowest (4-5) concentrations of the drugs).

```{r}
givePreatreatSum = function(predNum) {
  
  # get model coefficients and reshape
  coeffs <- do.call(cbind,lapply(predvar2[[predNum]], "[[", 'coeffs'))
  coeffs <- coeffs[-1,]
  coeffs <- as.matrix(coeffs)
  colnames(coeffs) <- unlist(drs["1:5"])
  colnames(coeffs) <- drugs[colnames(coeffs),"name"]
  coeffDF <- melt(as.matrix(coeffs))
  colnames(coeffDF) <- c("predictor", "drug", "beta")
  coeffDF$selected <- coeffDF$beta !=0
  
  #sort by times being selected
  coeffDF$predictor <- factor(coeffDF$predictor, level=)
  
  # number of drugs a predictor is chosen for
  gg1 <- coeffDF %>% group_by(predictor) %>% 
    dplyr::summarize(selectedSum = sum(selected)) %>%
    mutate(predictor = factor(predictor,
                              levels=predictor[order(selectedSum)])) %>%
    ggplot(aes(x=predictor, y=selectedSum)) + 
    geom_bar(stat="identity")+ylab("# drugs selected for") +
    coord_flip()
  
  # boxplots of non-zero coeffients
  orderMedian <- filter(coeffDF, selected) %>% group_by(predictor) %>% 
    dplyr::summarize(medianBeta = median(abs(beta)))
  coeffDF$predictor <- factor(
    coeffDF$predictor,
    levels=orderMedian$predictor[order(orderMedian$medianBeta)] )
  gg2 <- ggplot(filter(coeffDF, selected), aes(x=predictor, y=abs(beta))) +
    geom_boxplot() +
    coord_flip() + ggtitle("Distribution of non-zero coefficients")
  gridExtra::grid.arrange(gg1,gg2, ncol=1)
  
  # coefficeints per drug
  ggplot(filter(coeffDF, selected), 
         aes(x= drug, y=abs(beta), col= predictor=="Pretreatment")) +
    geom_point() +
    coord_flip()
  
  #drugs pretreatment is selected for
  as.character(filter(coeffDF, predictor=="Pretreatment" & beta!=0)$drug)
  PselDrugs <- as.character(
    filter(coeffDF, predictor=="Pretreatment" & beta!=0)$drug)
  length(PselDrugs)
  length(drs[[1]])
}
```

__Conc. 1-5__
```{r pretreatment_c1-5, echo=FALSE, fig.path=plotDir, dev=c("png", "pdf"), fig.width=7, fig.height=10, eval=TRUE}
givePreatreatSum(predNum=1)
```

__Conc. 4-5__
```{r pretreatment_c4-5, echo=FALSE, fig.path=plotDir, dev=c("png", "pdf"), fig.width=7, fig.height=10, eval=TRUE}
givePreatreatSum(predNum=2)
```



```{r, include=!exists(".standalone"), eval=!exists(".standalone")}
sessionInfo()
```

```{r, message=FALSE, warning=FALSE, include=FALSE}
rm(list=ls())
```