---
title: "Part 2"
output:
  BiocStyle::html_document
---

```{r, message=FALSE, include=!exists(".standalone"), eval=!exists(".standalone")}
library("BloodCancerMultiOmics2017")
library("reshape2") # melt
library("Biobase")
library("dplyr")
library("RColorBrewer")
library("ggplot2")
library("ggdendro")
library("gtable")
library("grid")
library("Rtsne")
library("ggbeeswarm")
```

```{r echo=FALSE}
plotDir = ifelse(exists(".standalone"), "", "part02/")
if(plotDir!="") if(!file.exists(plotDir)) dir.create(plotDir)
```

```{r}
options(stringsAsFactors=FALSE)
```

# Drug-induced effects on cell viability

Loading the data.
```{r}
data("lpdAll")
```

Here we show a relative cell viability (as compared to negative control) under treatment with 64 drugs at 5 concentrations steps each.

Prepare data.
```{r}
#select drug screening data on patient samples
lpd <- lpdAll[fData(lpdAll)$type == "viab" & pData(lpdAll)$Diagnosis != "hMNC",]
viabTab <- exprs(lpd)
viabTab <- viabTab[,complete.cases(t(viabTab))]
viabTab <- reshape2::melt(viabTab)
viabTab$Concentration <- fData(lpd)[viabTab$Var1,"subtype"]
viabTab <- viabTab[viabTab$Concentration %in% c("1","2","3","4","5"),]
viabTab$drugName <- fData(lpd)[viabTab$Var1,"name"]
viabTab <- viabTab[order(viabTab$Concentration),]

#order drug by mean viablitity
drugOrder <- group_by(viabTab, drugName) %>%
  summarise(meanViab = mean(value)) %>%
  arrange(meanViab)
viabTab$drugName <- factor(viabTab$drugName, levels = drugOrder$drugName)
```

Scatter plot for viabilities and using colors for concentrations.
```{r ViabilityScatter_main, echo=FALSE, fig.path=plotDir, dev=c("png", "pdf"), fig.width=10, fig.height=5, warning=FALSE, out.width=560, out.height=280}
#FIG# S1

#Color for each concentration
colorCode <- rev(brewer.pal(7,"Blues")[3:7])
names(colorCode) <- unique(viabTab$Concentration)

ggplot(viabTab, aes(x=drugName,y=value, color=Concentration)) +
  geom_jitter(size=1, na.rm = TRUE, alpha=0.8, shape =16) +
  scale_color_manual(values = colorCode) +
  ylab("Viability") + ylim(c(0,1.2)) + xlab("") +
  guides(color = guide_legend(override.aes = list(size=3,alpha=1),
                              title = "concentration index")) +
  theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5),
        legend.key = element_blank())
```

The plot shows high variability of effects between different drugs, from mostly lethal
(left) to mostly neutral (right), concentration dependence of effects and high variability of effects of the same drug/concentration across patients.

# Drug-drug correlation

We compared response patterns produced by drugs using phenotype clustering within diseases (CLL, T-PLL and MCL) using Pearson correlation analysis. The results imply that the drug assays probe tumor cells’ specific dependencies on survival pathways.

Loading the data.
```{r}
data("drpar", "patmeta", "drugs")
```

## Additional processing functions and parameters

Function that return the subset of samples for a given diagnosis (or all samples if diag=NA).
```{r}
givePatientID4Diag = function(pts, diag=NA) {
  pts = if(is.na(diag)) {
    names(pts)
  } else {
    names(pts[patmeta[pts,"Diagnosis"]==diag])
  }
  pts
}
```

Function that returns the viability matrix for given screen (for a given channel) for patients with given diagnosis.
```{r}
giveViabMatrix = function(diag, screen, chnnl) {
  data = if(screen=="main") drpar
          else print("Incorrect screen name.")
  pid = colnames(data)
  if(!is.na(diag))
    pid = pid[patmeta[pid,"Diagnosis"]==diag]
  
  return(assayData(data)[[chnnl]][,pid])
}
```

Function which reorders the clustered matrix based on computed singular value decomposition.
```{r}
callback = function(hc, mat){
   sv = svd(t(mat))$v[,1]
   dend = reorder(as.dendrogram(hc), wts = sv)
   rv <- as.hclust(dend)
   rv
}
```

Color scales for the heat maps.
```{r}
palette.cor1 = c(rev(brewer.pal(9, "Blues"))[1:8],
                 "white","white","white","white",brewer.pal(7, "Reds"))
palette.cor2 = c(rev(brewer.pal(9, "Blues"))[1:8],
                 "white","white","white","white",brewer.pal(7, "YlOrRd"))
```

## CLL/T-PLL

Pearson correlation coefficients were calculated based on the mean drug response for the two lowest concentration steps in the main screen and across CLL and T-PLL samples separately. Square correlation matrices were plotted together, with CLL in lower triangle and T-PLL in upper triangle. The drugs in a heat map are ordered by hierarchical clustering applied to drug responses of CLL samples.

```{r}
main.cll.tpll = BloodCancerMultiOmics2017:::makeCorrHeatmap(
  mt=giveViabMatrix(diag="CLL", screen="main", chnnl="viaraw.4_5"),
  mt2=giveViabMatrix(diag="T-PLL", screen="main", chnnl="viaraw.4_5"),
  colsc=palette.cor2, concNo="one")
```

```{r main.CLL.T-PLL, echo=FALSE, fig.path=plotDir, dev=c("png", "pdf"), fig.width=main.cll.tpll[["figure"]][["width"]], fig.height=main.cll.tpll[["figure"]][["height"]]}
#FIG# 2A
grid.draw(main.cll.tpll[["figure"]][["plot"]])
```

```{r main.cll.tpll.legend, echo=FALSE, fig.path=plotDir, dev=c("png", "pdf"), fig.width=main.cll.tpll[["legend"]][["width"]], fig.height=main.cll.tpll[["legend"]][["height"]], out.width=300, out.height=150}
#FIG# 2A
grid.draw(main.cll.tpll[["legend"]][["plot"]])
```

The major clusters in CLL include: kinase inhibitors targeting the B cell receptor, including idelalisib (PI3K), ibrutinib (BTK), duvelisib (PI3K), PRT062607 (SYK); inhibitors of redox signalling / reactive oxygen species (MIS−43, SD07, SD51); and BH3-mimetics (navitoclax, venetoclax).

### Effect of drugs with similar target

Here we compare the effect of drugs designed to target components of the same signalling pathway. 

```{r}
# select the data
mtcll = as.data.frame( t( giveViabMatrix(diag="CLL"  , screen="main", chnnl="viaraw.4_5") ) )
colnames(mtcll)  <- drugs[colnames(mtcll),"name"]

# function which plots the scatter plot
scatdr <-function(drug1, drug2, coldot, mt, min){

  p <- ggplot(data=mt,  aes(mt[,drug1],  mt[,drug2] ) )  + labs(x = drug1, y = drug2) +
        ylim(c(min, 1.35)) +  xlim(c(min, 1.35)) +
        theme(
           panel.background = element_blank(),
           axis.text = element_text(size = 15),
           axis.title = element_text(size = rel(1.5)),
           axis.line.x = element_line(colour = "black", size = 0.5),
           axis.line.y = element_line(colour = "black", size = 0.5),
           panel.grid.major = element_blank(),
           panel.grid.minor = element_blank() 
           ) +
       geom_point(size=3, col=coldot, alpha=0.8) +
       geom_text( aes(1, min+0.1, label = paste0("Pearson-R = ", round(cor( mt[,drug1],  mt[,drug2] ), 2) )), size = 5) +
       geom_smooth(method=lm) 

  return(p)
}
```


```{r cor_scatter, echo=FALSE, fig.path=plotDir, dev=c("png", "pdf"), , fig.width=6, fig.height=4, warning=FALSE, out.width=420, out.height=280}
#FIG# 2A
scatdr("ibrutinib", "spebrutinib", coldot="deeppink1", mt=mtcll, min=0.4)
scatdr("ibrutinib", "PRT062607 HCl", coldot="deeppink1", mt=mtcll, min=0.4)
scatdr("ibrutinib", "idelalisib", coldot="deeppink1", mt=mtcll, min=0.4)
scatdr("venetoclax", "navitoclax", coldot="goldenrod2", mt=mtcll, min=0.2)
scatdr("SD51", "MIS-43", coldot="dodgerblue3", mt=mtcll, min=0.2)
```


## T-PLL

Pearson correlation coefficients were calculated based on the mean drug response for the two lowest concentration steps in the main screen across T-PLL samples.

```{r}
main.tpll = BloodCancerMultiOmics2017:::makeCorrHeatmap(
  mt=giveViabMatrix(diag="T-PLL", screen="main", chnnl="viaraw.4_5"),
  colsc=palette.cor1, concNo="one")
```

```{r main.T-PLL, echo=FALSE, fig.path=plotDir, dev=c("png", "pdf"), fig.width=main.tpll[["figure"]][["width"]], fig.height=main.tpll[["figure"]][["height"]]}
#FIG# S6 B
grid.draw(main.tpll[["figure"]][["plot"]])
```

```{r main.tpll.legend, echo=FALSE, fig.path=plotDir, dev=c("png", "pdf"), fig.width=main.tpll[["legend"]][["width"]], fig.height=main.tpll[["legend"]][["height"]], out.width=300, out.height=150}
#FIG# S6 B
grid.draw(main.tpll[["legend"]][["plot"]])
```

Clusters of drugs with high correlation and anti-correlation are shown by red and blue squares, respectively.

Inhibitors of redox signaling / reactive oxygen species (MIS-43, SD07, SD51) are clustering together. Otherwise, in T-PLL samples correlations are not well pronounced.

## MCL

Pearson correlation coefficients were calculated based on the mean drug response for the two lowest concentration steps in the main screen across MCL samples.

```{r}
main.mcl = BloodCancerMultiOmics2017:::makeCorrHeatmap(
  mt=giveViabMatrix(diag="MCL", screen="main", chnnl="viaraw.4_5"),
  colsc=palette.cor1, concNo="one")
```

```{r main.MCL, echo=FALSE, fig.path=plotDir, dev=c("png", "pdf"), fig.width=main.mcl[["figure"]][["width"]], fig.height=main.mcl[["figure"]][["height"]]}
#FIG# S6 A
grid.draw(main.mcl[["figure"]][["plot"]])
```

```{r main.mcl.legend, echo=FALSE, fig.path=plotDir, dev=c("png", "pdf"), fig.width=main.mcl[["legend"]][["width"]], fig.height=main.mcl[["legend"]][["height"]], out.width=300, out.height=150}
#FIG# S6 A
grid.draw(main.mcl[["legend"]][["plot"]])
```

Clusters of drugs with high correlation and anti-correlation are shown by red and blue squares, respectively.

The major clusters include: kinase inhibitors of the B cell receptor, incl. idelalisib (PI3K), ibrutinib (BTK), duvelisib (PI3K), PRT062607 (SYK); inhibitors of redox signaling / reactive oxygen species (MIS-43, SD07, SD51) and BH3 mimetics (navitoclax, venetoclax).


# Disease-specific drug response phenotypes

Loading the data.
```{r}
data(list=c("lpdAll", "conctab", "patmeta"))
```

Preprocessing of drug screen data.
```{r}
#Select rows contain drug response data
lpdSub <- lpdAll[fData(lpdAll)$type == "viab",]

#Only use samples with complete values
lpdSub <- lpdSub[,complete.cases(t(exprs(lpdSub)))]

#Transformation of the values
exprs(lpdSub) <- log(exprs(lpdSub))
exprs(lpdSub) <- t(scale(t(exprs(lpdSub))))

#annotation for drug ID
anno <- sprintf("%s(%s)",fData(lpdSub)$name,fData(lpdSub)$subtype)
names(anno) <- rownames(lpdSub)
```

Function to run t-SNE.
```{r}
tsneRun <- function(distMat,perplexity=10,theta=0,max_iter=5000, seed = 1000) {
  set.seed(seed)
  tsneRes <- Rtsne(distMat, perplexity = perplexity, theta = theta, 
                   max_iter = max_iter, is_distance = TRUE, dims =2)
  tsneRes <- tsneRes$Y
  rownames(tsneRes) <- labels(distMat)
  colnames(tsneRes) <- c("x","y")
  tsneRes
}
```

Setting color scheme for the plot.
```{r}
colDiagFill = c(`CLL` = "grey80",
    `U-CLL` = "grey80",
    `B-PLL`="grey80",
    `T-PLL`="#cc5352",
    `Sezary`="#cc5352",
    `PTCL-NOS`="#cc5352",
    `HCL`="#b29441",
    `HCL-V`="mediumaquamarine",
    `AML`="#addbaf",
    `MCL`="#8e65ca",
    `MZL`="#c95e9e",
    `FL`="darkorchid4",
    `LPL`="#6295cd",
    `hMNC`="pink")

colDiagBorder <- colDiagFill
colDiagBorder["U-CLL"] <- "black"
```

Sample annotation.
```{r}
annoDiagNew <- function(patList, lpdObj = lpdSub) {
  Diagnosis <- pData(lpdObj)[patList,c("Diagnosis","IGHV Uppsala U/M")]
  DiagNew <- c()
  
  for (i in seq(1:nrow(Diagnosis))) {
   if (Diagnosis[i,1] == "CLL") {
      if (is.na(Diagnosis[i,2])) {
        DiagNew <- c(DiagNew,"CLL")
      } else if (Diagnosis[i,2] == "U") {
        DiagNew <- c(DiagNew,sprintf("%s-%s",Diagnosis[i,2],Diagnosis[i,1]))
      } else if (Diagnosis[i,2] == "M") {
        DiagNew <- c(DiagNew,"CLL")
      }
    } else DiagNew <- c(DiagNew,Diagnosis[i,1])
  }
  DiagNew
}
```

Calculate t-SNE and prepare data for plotting the result.
```{r}
#prepare distance matrix
distLpd <- dist(t(exprs(lpdSub)))

#run t-SNE
plotTab <- data.frame(tsneRun(distLpd,perplexity=25, max_iter=5000, seed=338))

#annotated patient sample
plotTab$Diagnosis <- pData(lpdSub[,rownames(plotTab)])$Diagnosis
plotTab$Diagnosis <- annoDiagNew(rownames(plotTab,lpdSub)) #consider IGHV status
plotTab$Diagnosis <- factor(plotTab$Diagnosis,levels = names(colDiagFill))
```

```{r tSNE, echo=FALSE, fig.path=plotDir, dev=c("png", "pdf"), fig.width=10, fig.height=8, warning=FALSE, out.width=600, out.height=480}
#FIG# 2 C
p <- (ggplot(plotTab, aes(x=x,y=y)) +
        geom_point(size=3, shape= 21, aes(col = Diagnosis, fill = Diagnosis)) +
        theme_classic() +
        theme(axis.ticks=element_line(color="black",size=0.5),
              text=element_text(size=20),
              axis.line.x = element_line(color="black",size=0.5),
              axis.line.y = element_line(color="black",size=0.5),
              legend.position="right") +
        scale_fill_manual(values = colDiagFill) +
        scale_color_manual(values = colDiagBorder) +
        xlab("Component 1") + ylab("Component 2")) +
  coord_cartesian(xlim = c(-20,20),ylim=c(-20,20))

print(p)
```

## Example: dose-response curves

Here we show dose-response curve for selected drugs and patients.

First, change concentration index into real concentrations according to `conctab`.
```{r}
lpdPlot <- lpdAll[fData(lpdAll)$type == "viab",]
concList <- c()
for (drugID in rownames(fData(lpdPlot))) {
  concIndex <- as.character(fData(lpdPlot)[drugID,"subtype"])
  concSplit <- unlist(strsplit(as.character(concIndex),":"))
  ID <- substr(drugID,1,5)
  if (length(concSplit) == 1) {
    realConc <- conctab[ID,as.integer(concSplit)]
    concList <- c(concList,realConc)
  } else {
    realConc <- sprintf("%s:%s",
                        conctab[ID,as.integer(concSplit[1])],
                        conctab[ID,as.integer(concSplit[2])])
    concList <- c(concList,realConc)
  }
}

fData(lpdPlot)$concValue <- concList
lpdPlot <- lpdPlot[,complete.cases(t(exprs(lpdPlot)))]
```

Select drugs and samples.
```{r}
patDiag <- c("CLL","T-PLL","HCL","MCL")
drugID <- c("D_012_5","D_017_4","D_039_3","D_040_5","D_081_4","D_083_5")

lpdBee <- lpdPlot[drugID,pData(lpdPlot)$Diagnosis %in% patDiag]
```

Prepare the data for plot
```{r}
lpdCurve <-
  lpdPlot[fData(lpdPlot)$name %in% fData(lpdBee)$name,
          pData(lpdPlot)$Diagnosis %in% patDiag]
lpdCurve <- lpdCurve[fData(lpdCurve)$subtype %in% seq(1,5),]
dataCurve <- data.frame(exprs(lpdCurve))
dataCurve <- cbind(dataCurve,fData(lpdCurve)[,c("name","concValue")])
tabCurve <- melt(dataCurve,
                 id.vars = c("name","concValue"), variable.name = "patID")
tabCurve$Diagnosis <- factor(pData(lpdCurve[,tabCurve$patID])$Diagnosis,
                             levels = patDiag)
tabCurve$value <- tabCurve$value
tabCurve$concValue <- as.numeric(tabCurve$concValue)

# set order
tabCurve$name <- factor(tabCurve$name, levels = fData(lpdBee)$name)

#calculate the mean and mse for each drug+cencentration in different disease
tabGroup <- group_by(tabCurve,name,concValue,Diagnosis)
tabSum <- summarise(tabGroup,meanViab = mean(value))
```


Finally, plot dose-response curve for each selected drug.
```{r viabilityCurve, fig.path=plotDir, dev=c("png", "pdf"), fig.width=4, fig.height=3}
#FIG# 2 C
tconc = expression("Concentration [" * mu * "M]")
fmt_dcimals <- function(decimals=0){
   # return a function responpsible for formatting the 
   # axis labels with a given number of decimals 
   function(x) as.character(round(x,decimals))
}

for (drugName in unique(tabSum$name)) {
  tabDrug <- filter(tabSum, name == drugName)
  p <- (ggplot(data=tabDrug, aes(x=concValue,y=meanViab, col=Diagnosis)) +
          geom_line() + geom_point(pch=16, size=4) +
          scale_color_manual(values = colDiagFill[patDiag])
        + theme_classic() +
          theme(panel.border=element_blank(),
                axis.line.x=element_line(size=0.5,
                                         linetype="solid", colour="black"),
                axis.line.y = element_line(size = 0.5,
                                           linetype="solid", colour="black"),
                legend.position="none",
                plot.title = element_text(hjust = 0.5, size=20),
                axis.text = element_text(size=15),
                axis.title = element_text(size=20)) +
          ylab("Viability") + xlab(tconc) + ggtitle(drugName) +
          scale_x_log10(labels=fmt_dcimals(2)) +
          scale_y_continuous(limits = c(0,1.3), breaks = seq(0,1.3,0.2)))
  plot(p)
}
```

## Example: drug effects as bee swarms

```{r viabilityBee, fig.path=plotDir, dev=c("png", "pdf"), fig.width=5, fig.height=10, warning=FALSE}
#FIG# 2 D
lpdDiag <- lpdAll[,pData(lpdAll)$Diagnosis %in% c("CLL", "MCL", "HCL", "T-PLL")]
dr <- c("D_012_5", "D_083_5", "D_081_3", "D_040_4", "D_039_3")

m <- data.frame(t(exprs(lpdDiag)[dr, ]), diag=pData(lpdDiag)$Diagnosis)
m <- melt(m)
m$lable <- 1
for (i in 1:nrow(m )) {
  m[i, "lable"] <- giveDrugLabel(as.character(m[i, "variable"]), conctab, drugs)
} 

  
ggplot( m, aes(diag, value, color=factor(diag) ) ) +
  ylim(0,1.3) + ylab("Viability") + 
  xlab("") +
  geom_boxplot(outlier.shape = NA) +
  geom_beeswarm(cex=1.4, size=1.4,alpha=0.5, color="grey80") +
  scale_color_manual("diagnosis", values=c(colDiagFill["CLL"], colDiagFill["MCL"], 
                                           colDiagFill["HCL"], colDiagFill["T-PLL"])) +
  theme_bw() +
  theme(legend.position="right") +
  theme(
    panel.background =  element_blank(), 
    panel.grid.minor.x =  element_blank(),
    axis.text = element_text(size=15),
    axis.title = element_text(size=15),
    strip.text = element_text(size=15)
  ) +
  facet_wrap(~ lable, ncol=1) 
```


```{r, include=!exists(".standalone"), eval=!exists(".standalone")}
sessionInfo()
```

```{r, message=FALSE, warning=FALSE, include=FALSE}
rm(list=ls())
```
